Supporting Awareness of Indirect Conflicts with Impact

### 2013-06-05 impact-position-2013/paper.tex
Added new version of impact position paper.

% Paper abstract
\begin{abstract}
Awareness has been largely studied in the field of CSCW research within
software engineering. Many tools and techniques have been proposed and built
in order to provide software developers and stakeholders with a greater
sense of workspace and task awareness within their software projects. These
techniques and tools have been largely focused on detecting \textit{direct} 
conflicts which arise over a project's life time or have created 
an exploratory ground for stakeholders to use as a means of resolving self
discovered direct or indirect conflicts. However, detecting 
and providing pertinent information regarding \textit{indirect} conflicts has been
largely ignored partially due to their inherently larger
complexity than direct conflicts. Indirect conflicts arise when changes
in one software artifact affect another. In this paper, we present
\textit{Impact}, a new task awareness tool directly aimed at both detecting
and presenting indirect conflicts which arise inside of a software project.
\textit{Impact} represents a first step towards the design and implementation
of awareness tools that specifically address indirect conflicts in software
projects. To describe \textit{Impact}, we introduce previous indirect 
conflict awareness attempts, our design 
and implementation, as well as \textit{Impact's} potential through
two evaluation studies. 
\end{abstract}

@@

Indirect conflicts arising in source code are inherently
difficult to resolve as most of the time, source code analysis must
be performed in order to find relationships between technical objects
which are harmed by changes.
While some awareness tools have been created with these indirect conflicts
primarily in mind~\cite{Begel:2010:CDE, Trainer:2005:BGT}, most have only 
created an exploratory environment which is used by developers to
solve conflicts which may arise. These tools were not designed to detect
indirect conflicts that arise and alert developers to their presence 
inside the software system. Sarma et al.~\cite{Sarma:2007:TSA} has started to work directly
on solving indirect conflicts, however, these products are not designed to handle
internal structures of technical objects.

In this paper, we report on our research into supporting indirect conflicts
and present the design, implementation, and evaluation of the tool \textit{Impact},
a web based tool that aims at detecting indirect conflicts among developers
and notifying the appropriate members involved in these conflicts.
By leveraging technical relationships inherent of 
software projects with method call graphs~\cite{Lakhotia:1993:CCM}
as well as detecting changes
to these technical relationship through software configuration management
(SCM) systems, \textit{Impact} is able to detect indirect conflicts as well as
alert developers involved in such conflicts in task awareness. While this paper
outlines \textit{Impact's} specific implementation, its design is rather
generic and can be implemented in similar indirect conflict awareness tools.
\textit{Impact} represents a first step towards the design and implementation
of awareness tools which address indirect conflicts in software development.

The rest of this paper is organized as follows. First, we begin by discussing
similar indirect conflict awareness tools which have partially solved the
issues presented by this paper and how their designs can be applied to 
\textit{Impact}. In the following section we describe a generic design and implementation
of \textit{Impact} as an awareness tool. We then discuss a preliminary evaluation of
\textit{Impact} followed by conclusions and future work.

\section{Related Work}
Although there is an abundance of awareness tools developed in research
today, only a handful have made an attempt to examine indirect conflicts.
Here, we will outline three of the forefront projects in indirect conflicts
and how these projects have influenced the decision making process in
the design and implementation of \textit{Impact}.

We first start with both Codebook~\cite{Begel:2010:CDE} and 
Ariadne~\cite{Trainer:2005:BGT}. These projects produce an exploratory
environment for developers to handle indirect conflicts. Exploratory
pertains to the ability to solve self determined conflicts, meaning that
once a developer discovers a conflict, they can use the tool as a type of
lookup table to solve their issue. Codebook is a type of social developer
network that relates developers to source code, issue repositories and
other social media while Ariadne only looks at source code for developer
to source code association. Through Codebook, developers become
owners of source code artifacts. Both projects also use program 
dependency graphs~\cite{Horwitz:1992:UPD}
in order to relate technical artifacts to each other. These projects make 
use of method call graphs in order to 
determine which methods invoke others which forms the basis for 
linking source code artifacts creating a directed graph. While these 
projects can be great tools 
for solving indirect conflicts which may arise, by querying such directed
graphs to view impacts of conflict creating code, they lack the ability to
detect conflicts on their own.

@@

\textit{Impact} is designed to address these limitations by extending this work in
two directions: (1) the detection of indirect conflicts at an internal level
of technical objects as opposed from object signatures, and (2) the dissemination of
awareness information to all appropriate developers regardless of their current
workspace activities. \textit{Impact} is also designed around the successes these
projects have had in the past with directed graphs as 
well as elements of collection, ownership, and distribution functionality.

@@

While technical object relationships form the basis of indirect conflicts,
communication between developers is the ultimate goal of resolving such conflicts.
This being the case, developer ownership must be placed on the 
identified technical objects. With this ownership, we now infer
relationships among developers based on their technical objects
``uses'' relationship. Developer A, who owns object 1, which uses 
object 2 owned by developer B, may be notified by a change to
object 2's internal workings. Most, if not all, ownership information
of technical objects can be extracted from a project's source code
repository (CVS, Git, SVN, etc.).

Finally, the indirect conflict tool must be able to detect changes
to the technical objects defined above and notify the appropriate owners
to the conflict. 
Two approaches have been 
proposed for change gathering techniques: real time and commit 
time~\cite{Fitzpatrick:2002:SPA}.
We propose the use of commit time
information gathering as it avoids the issue of developers 
overwriting previous work or deleting modifications which would 
produce information for changes that no longer exist. However, the
trade off is that indirect conflicts must be committed before detected,
which results in conflicts being apart of the system before being able
to be dealt with as opposed to catching conflicts before they happen.
At commit time, the tool must parse changed source code in relation
to technical artifacts in the created directed graph detailed above.
Where \textit{Impact's} design differs from that of Palantir's is that
the object's entire body (method definition and internal body) is 
parsed at commit time to detect changes anywhere in the technical object.
Once technical objects are found to be changed, appropriate owners
of objects which use the changed object should be notified.
In Figure~\ref{fig:graph}, Carl changes method (technical object) 1,
which effects methods 2 and 3 resulting in the alerting of
developers Bret, Daniel, and Anne.

With this three step design: (i) creating directed graphs of technical
objects, (ii) assigning ownership to those technical objects, and (iii)
detecting changes within commit time and the notification of conflict information
to appropriate owners, we believe a wide variety of
indirect conflict awareness tools can be created or extended. The
implementation of \textit{Impact} described in the following section follows these
three design guidelines.

\subsection{Implementation}
For \textit{Impact's} implementation, we decided to focus on methods as our
selected technical objects to infer a directed graph from. The ``uses'' 
relationship described above for methods is method invocation.
Thus, in our constructed directed graph, methods represent nodes
and method invocations represent our directed edges. In order to 
construct this directed graph, abstract syntax trees (ASTs) are 
constructed from source files in the project. The ASTs allow us
to construct method call graphs from which the directed 
graphs can be constructed.

Once the directed graph is constructed, we must now assign
ownership to our technical objects (methods) as per our design.
To do this, we simply query the source code repository. In our case
we used Git as the source code repository, so the command \textit{git blame}
is used for querying ownership information. (Most source code 
repositories have similar commands and functionality.) This command 
returns the source code authors per linee which can is used to assign
ownership to methods. If a method has 10 lines and developer A
has written 3 while developer B has written 7, then ownership is
assigned 30\% and 70\% respectively.

To detect changes to our technical objects (methods), we simply 
use a commit's \textit{diff} which is a representation of all changes
made inside a commit. We can use the lines changed in the \textit{diff} to 
find methods that have been changed. This gives cause of potential
indirect conflicts. We now find all methods in our directed graphs
which invoke these changed methods. These are the final indirect
conflicts.

Once the indirect conflicts have been found, we use the
ownership information of our technical objects to send notifications to
those developers involved in the indirect conflict. All owners
of methods which invoke those that have been changed are alerted
to the newly changed method. This can been seen in
Figure~\ref{fig:impact}, the user interface of \textit{Impact}. Here, in an RSS type
feed, the changing developer, time of change, changed method,
invoking methods, and commit message are all displayed. This 
list is auto updating as new commits are pushed to the central
git repository. The information provided also comes with a simple
weight of the percent changed of changed method multiplied by 
ownership of the invoking method. This allows developers to filter
through high and low changes affecting their own source code.
Developers are now able to be alerted of detected indirect
conflicts and solve them by communicating the developer
who authored the change.

@@

\section{Evaluation}
To fully evaluate both the generic design of detecting and resolving
indirect conflicts as well as \textit{Impact}, extensive testing and evaluation
must be performed. However, we feel that a simple evaluation is
first needed to assess the foundation of \textit{Impact}'s design and claims
about indirect conflicts at the method level.

We performed a user case study where we gave \textit{Impact} to 
two small development teams composed of three developers. Each team was
free to use \textit{Impact} at their leisure during their development process,
after which interviews were conducted with lead developers from 
each development team. The interviews were conducted after each
team had used \textit{Impact} for three weeks.

We asked lead developers to address two main concerns: do indirect
conflicts pose a threat
at the method level (e.g. method 1 has a bug because it invokes method
2 which has has its implementation changed recently), 
and did \textit{Impact} help raise
awareness and promote quicker conflict resolution for indirect
conflicts. Our two interviews largely supported our expectation of
indirect conflicts posing a serious threat to developers, especially
in medium to large teams or projects as opposed to the small
teams which they were apart of. It was also pointed
out that method use can be a particularly large area for indirect
conflicts to arise. However, it was noted that
any technical object which is used as an interface to some data
construct or methodology, database access for instance, can be 
a large potential issue for indirect conflicts.  Interview response to
\textit{Impact} was also largely positive, as interviewees stated that \textit{Impact}
helped raise awareness among their teams with what other developers
are doing as well as the influence it has on their own work. However,
\textit{Impact} was shown to have information overload. It was suggested
that while all method changes were being detected,
not all are notification worthy. One developer suggested to only notify
developers to indirect conflicts if the internal structure of a method
changes due to modification to input parameters or output parameters.
In other words, the boundaries of the technical objects (changing
how a parameter is used inside the method, modifying the return
result inside the method) seem to be more of interest than other 
internal workings. More complex inner workings of methods were also noted 
to be of interest to developers
such as cyclomatic complexity, or time and space requirements.

These two studies have shown that our design and approach to
detecting and alerting developers to indirect conflicts appear
to be on the correct path. \textit{Impact} as a tool has laid the foundations
for future work in detecting indirect conflicts as well as notifying
developers, although more thought must be given as to 
what constitutes a meaningful change inside our selected 
technical objects to avoid issues of information overload.

\section{Next Steps in Research of Indirect Conflicts}
While \textit{Impact} has laid the foundations for awareness tool construction
aimed at resolving indirect conflicts, it has also raised questions 
in regards to levels of granularity in notification of indirect conflicts and the
implied trade off between notifications and information overload. To address
these problems, we plan to find industry expert opinion through interviews
and surveys with experienced developers. From these interactions, we plan to
answer some of the questions presented in this paper such as which technical
objects concern developers with indirect conflicts, and what types of changes
to these objects are cause for notification. We are also planning, through these
expert opinions, to further the 
need for indirect conflict awareness tools by establishing that indirect
conflicts are real world problems with harmful consequences. 
Further more, we plan to bring these opinions back to the realm of 
\textit{Impact}, in combination with other academic works on balancing
between notifications and information overload~\cite{Wang:2007}, to
continue our development
on this tool to further its life both as a research tool and a possible
industry application.

\section{Conclusion and Future Work}
In this paper, we have presented the issues that arise from indirect 
conflicts in present awareness tools. We have proposed a generic 
design for the future development of awareness tools in regards to
handling indirect conflicts. We have presented a prototype 
awareness tool, \textit{Impact}, which was designed around our generic 
technical object approach. \textit{Impact} was evaluated on a small scale, showing
its future potential as well as highlighting its current weaknesses. 
Our preliminary evaluation has given us concrete insights into 
understanding awareness tools that specifically address indirect conflicts and has
strengthened the need for further research towards notification and information
overload (what makes a change notification
worthy), which has become the goal of our next step.

>>>>

% Paper abstract
\begin{abstract}
Awareness has been largely studied in the field of CSCW research within
software engineering. Many tools and techniques have been proposed and built
in order to provide software developers and stakeholders with a greater
sense of workspace and task awareness within their software projects. These
techniques and tools have been largely focused on detecting \textit{direct} 
conflicts which arise over a project's life time or have created 
an exploratory ground for stakeholders to use as a means of resolving self
discovered direct or indirect conflicts. However, detecting 
and providing pertinent information regarding \textit{indirect} conflicts has been
largely ignored partially due to their inherently larger
complexity than direct conflicts. Indirect conflicts arise when changes
in one software artifact affect another. In this paper, we present
\textit{Impact}, a new task awareness tool directly aimed at both detecting
and presenting indirect conflicts which arise inside of a software project.
\textit{Impact} represents a first step towards the design and implementation
of awareness tools that specifically address indirect conflicts in software
projects. To describe \textit{Impact}, we introduce previous indirect 
conflict awareness attempts, the existing problems in the field, our design 
and implementation for a potential solution, as well as \textit{Impact's} future work
on a proposed solution to information overload which arises with indirect
conflict awareness.
\end{abstract}

@@

Indirect conflicts arising in source code are inherently
difficult to resolve as most of the time, source code analysis or
program slicing~\cite{Weiser:1981:PS} must
be performed in order to find relationships between technical objects
which are harmed by changes.
While some awareness tools have been created with these indirect conflicts
primarily in mind~\cite{Begel:2010:CDE, Trainer:2005:BGT}, most have only 
created an exploratory environment which is used by developers to
solve conflicts which may arise. These tools were not designed to detect
indirect conflicts that arise and alert developers to their presence 
inside the software system. Sarma et al.~\cite{Sarma:2007:TSA} has started to work directly
on solving indirect conflicts, however, these products are not designed to handle
internal structures of technical objects.

While indirect conflicts remain to be a large problem area in task awareness,
some attempts have been made to both detect and provide developer awareness
of indirect conflicts. Servant et al.~\cite{Servant:2010:CPI}, have devised a tool
which can both detect and alert developers to indirect conflicts arising in 
their software projects. However, a main issue of information overload continues
to arise from this and many other indirect conflict tools. As software systems
grow larger and more complex, indirect dependencies tend to become more numerous.
This causes small changes in source code to have a large impact on the software system
which these tools often report exhaustively. Information overload must be addressed
in any feasible attempt made on indirect conflicts.

In this paper, we report on our research into supporting indirect conflicts
and present the design, implementation, and future evaluation of the tool \textit{Impact},
a web based tool that aims at detecting indirect conflicts among developers
and notifying the appropriate members involved in these conflicts.
By leveraging technical relationships inherent of 
software projects with method call graphs~\cite{Lakhotia:1993:CCM}
as well as detecting changes
to these technical relationship through software configuration management
(SCM) systems, \textit{Impact} is able to detect indirect conflicts as well as
alert developers involved in such conflicts in task awareness while limiting information
overload by using design by contract~\cite{} solutions to method design. While this paper
outlines \textit{Impact's} specific implementation, its design is rather
generic and can be implemented in similar indirect conflict awareness tools.
\textit{Impact} represents a first step towards the design and implementation
of awareness tools which address indirect conflicts in software development.

The rest of this paper is organized as follows. First, we begin by discussing
similar indirect conflict awareness tools which have partially solved the
issues presented by this paper and how their designs can be applied to 
\textit{Impact}. In the following section we describe a generic design and implementation
of \textit{Impact} as an awareness tool. We then discuss future work in terms
of evaluation and further understanding of the problems at hand.

\section{Related Work}
Although there is an abundance of awareness tools developed in research
today, only a handful have made an attempt to examine indirect conflicts.
Here, we will outline four of the forefront projects in indirect conflicts
and how these projects have influenced the decision making process in
the design and implementation of \textit{Impact}.

We first start with both Codebook~\cite{Begel:2010:CDE} and 
Ariadne~\cite{Trainer:2005:BGT}. These projects produce an exploratory
environment for developers to handle indirect conflicts. Exploratory
pertains to the ability to solve self determined conflicts, meaning that
once a developer discovers a conflict, they can use the tool as a type of
lookup table to solve their issue. Codebook is a type of social developer
network that relates developers to source code, issue repositories and
other social media while Ariadne only looks at source code for developer
to source code association. Through Codebook, developers become
owners of source code artifacts. Both projects also use program 
dependency graphs~\cite{Horwitz:1992:UPD}
in order to relate technical artifacts to each other. These projects make 
use of method call graphs in order to 
determine which methods invoke others which forms the basis for 
linking source code artifacts creating a directed graph. While these 
projects can be great tools 
for solving indirect conflicts which may arise, by querying such directed
graphs to view impacts of conflict creating code, they lack the ability to
detect potential conflicts on their own.

@@

We will lastly examine the tool CASI~\cite{Servant:2010:CPI} which uses
a sphere of influence for each developer to determine how source code
changes are indirectly related to other components of a software project.
CASI uses dependency slicing~\cite{Bajracharya:2009:SIS} instead of the 
call graphs in Ariadne~\cite{Trainer:2005:BGT} which gives dependencies among
all source code entities. This provides a verbose output of dependencies when
source code is changed. CASI also implements a visualization where
a developer can see what parts of a software projects he or she may be effecting
with the source code change. This allows the developer themselves to go and fix
potential issues elsewhere in the project before the code change is committed
to the software repository. While CASI covers great ground in its approach,
it still leaves the issue of information overload, although attempts were made
to solve this by having severity levels of indirect conflicts presented to
the user.

\textit{Impact} is designed to address these limitations by extending this work in
two directions: (1) the detection of high importance indirect conflicts at an internal level
of technical objects, (2) the dissemination of
awareness information to all appropriate developers regardless of their current
workspace activities, and (3) potential solutions to information overload. 
\textit{Impact} is also designed around the successes these
projects have had in the past with dependency graphs as 
well as elements of collection, ownership, and distribution functionality.

@@

While technical object relationships form the basis of indirect conflicts,
communication between developers is our ultimate goal of resolving such conflicts.
This being the case, developer ownership must be placed on the 
identified technical objects. With this ownership, we now infer
relationships among developers based on their technical objects
``uses'' relationship. Developer A, who owns object 1, which uses 
object 2 owned by developer B, may be notified by a change to
object 2's internal workings. Most, if not all, ownership information
of technical objects can be extracted from a project's source code
repository (CVS, Git, SVN, etc.).

Finally, the indirect conflict tool must be able to detect changes
to the technical objects defined above and notify the appropriate owners
to the conflict. 
Two approaches have been 
proposed for change gathering techniques: real time and commit 
time~\cite{Fitzpatrick:2002:SPA}.
We propose the use of commit time
information gathering as it avoids the issue of developers 
overwriting previous work or deleting modifications which would 
produce information for changes that no longer exist. However, the
trade off is that indirect conflicts must be committed before detected,
which results in conflicts being apart of the system before being able
to be dealt with as opposed to catching conflicts before they happen.
At commit time, the tool must parse changed source code in relation
to technical artifacts in the created directed graph detailed above.
Where \textit{Impact's} design differs from that of Palantir's is that
the object's entire body (method definition and internal body) is 
parsed, similar to that of CASI~\cite{Servant:2010:CPI},
at commit time, as opposed to real time, to detect changes anywhere in the technical object.
This is a first design step towards avoiding information overload.
Once technical objects are found to be changed, appropriate owners
of objects which use the changed object should be notified. However, to
avoid information overload at this step, analyses of the changed object
must be preformed to determine whether or not it is a ``notification worthy''
change. Our proposed solution to this is explained in Section~\ref{sec:ns}.
In Figure~\ref{fig:graph}, Carl changes method (technical object) 1,
which effects methods 2 and 3 resulting in the alerting of
developers Bret, Daniel, and Anne. We have opted to alert the invoking
developers rather than the developer making the change to potential
solutions as our conflicts are detected at commit time and this supports
the idea of a socio-technical congruence~\cite{Kwan:2011:ESC} 
from software structure to communication patterns in awareness systems.

With this three step design: (i) creating directed graphs of technical
objects, (ii) assigning ownership to those technical objects, and (iii)
detecting ``notification worthy'' changes at commit time and the dissemination of conflict information
to appropriate owners, we believe a wide variety of
indirect conflict awareness tools can be created or extended. The
on going implementation of \textit{Impact} described in the following 
section follows these three design guidelines.

\subsection{Implementation}
For \textit{Impact's} implementation, we decided to focus on methods as our
selected technical objects to infer a directed graph from. The ``uses'' 
relationship described above for methods is method invocation.
Thus, in our constructed dependency graph, methods represent nodes
and method invocations represent our directed edges. In order to 
construct this directed graph, abstract syntax trees (ASTs) are 
constructed from source files in the project.

Once the directed graph is constructed, we must now assign
ownership to our technical objects (methods) as per our design.
To do this, we simply query the source code repository. In our case
we used Git as the source code repository, so the command \textit{git blame}
is used for querying ownership information. (Most source code 
repositories have similar commands and functionality.) This command 
returns the source code authors per line which can be used to assign
ownership to methods.

To detect changes to our technical objects (methods), we simply 
use a commit's \textit{diff} which is a representation of all changes
made inside a commit. We can use the lines changed in the \textit{diff} to 
find methods that have been changed. This gives cause of potential
indirect conflicts. Here is where information overload can occur and must
be dealt with. Impact's current state does not address this as our potential 
solution is still under investigation, but is the
main focus of our next steps as detailed in Section~\ref{sec:ns}.
We now find all methods in our directed graphs which invoke these changed methods. 
These are the final indirect conflicts.

Once the indirect conflicts have been found, we use the
ownership information of our technical objects to send notifications to
those developers involved in the indirect conflict. All owners
of methods which invoke those that have been changed are alerted
to the newly changed method. Impact can been seen in
Figure~\ref{fig:impact}, the user interface of \textit{Impact}. Here, in an RSS type
feed, the changing developer, time of change, changed method,
invoking methods, and commit message are all displayed. 
The weight provided will be an indication of importance based
on the future work to be outline in Section~\ref{sec:ns}. This
again helps solve the issue of information overload.

@@

\section{Proposed Solution to Information Overload}
\label{sec:ns}
As was previously stated, the main issue to be addressed with Impact is
information overload. Our potential solution comes from the ideas of
Myers~\cite{} on ``design by contract''. In this methodology, changes to method
preconditions and postconditions are considered to be the most harmful. 
This includes adding conditions that must be met by both input
and output parameters such as limitations to input and expected
output. To achieve this level of source code analysis, the ideas of
Fluri et al.~\cite{Fluri:2007:CDT} can be used on the previously generated
ASTs for high granularity
source code change extraction when determining if preconditions or
postconditions have changed.
We plan to bring these ideas into the realm of what makes a
change ``notification worthy'', as to deal with information overload for
developers using Impact. By only examining ``notification worthy'' changes,
Impact would cut down the number of notifications 
issued to just those which developers care about. To investigate this solution, we plan to
interview and survey industry software developers at a wide range of
experience to determine what makes a change in software important. We believe
by only monitoring what developers and Myers~\cite{} find important to
indirect conflict awareness, we will be able to eliminate information overload
in Impact and other dependency awareness tools. We believe that these preconditions
and postconditions will serve as the main tool for dealing with information overload.

\section{Conclusion and Future Work}
In this paper, we have presented the issues that arise from indirect 
conflicts in present awareness tools. We have proposed a generic 
design for the future development of awareness tools in regards to
handling indirect conflicts. We have presented a prototype 
awareness tool, \textit{Impact}, which was designed around our generic 
technical object approach. We have also proposed a solution to information
overload which occurs in most indirect conflict awareness tools. We plan
on investigating our solution further through developer interview and
surveys, and by finally implementing the solution into Impact.
